Re-examining cross-cultural similarity judgments using language statistics

### 2023-02-02 code/4_paper/manuscript/manuscript.Rmd
done!

---
title: "Re-examining cross-cultural similarity judgments using language statistics"
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage[T1, T2A, T5]{fontenc}
  - \usepackage{kotex}
  - \usepackage{CJKutf8}
author-information: > 
    \author{{\large \bf Khuyen N. Le (knl005@ucsd.edu)}$^{1}$, {\large \bf Alexandra Carstensen (abc@ucsd.edu)}$^{1}$, \\ {\large \bf Shan Gao (shangaocog@gmail.com)$^2$}, {\large \bf Michael C. Frank (mcfrank@stanford.edu)$^3$},  \AND
    $^1$Department of Psychology, University of California, San Diego, $^2$Department of Psychology, University of Chicago, \\ $^3$Department of Psychology, Stanford University
    }
abstract: >
 Is “cow” more closely related to “grass” or to “chicken”? Speakers of different languages judge similarity in this context differently, but why? One possibility is that cultures co-varying with these languages induce differences in conceptualizations of similarity. Specifically, East Asian cultures may promote reasoning about thematic similarity, by which cow and grass are more related, whereas Western cultures may bias judgments toward taxonomic relations, like cow and chicken. We measure similarity judgments across the US, China, and Vietnam and replicate US-China cultural differences with Vietnam intermediate. We also show that word embedding models (fastText models for each language) are related to judgments within each country, suggesting a possible alternative interpretation of cross cultural differences. Perhaps notions of similarity are similar across contexts, but the statistics of the linguistic environment vary.

@@

Our study uses cosine distances of fastText word vectors as a measure of lexical co-occurrence.[^1]
fastText is a system that uses lexical co-occurrence information to generate a vector representing each word in its lexicon [@Mikolov2018].
Practically speaking, fastText vectors are available for many languages, and fastText has also been shown to be sensitive to cultural differences in word meanings: @Thompson2020 demonstrated that the distribution of semantic meaning clusters generated by fastText trained on language-specific corpora correlates with the cultural, historical, and geographical similarities of these languages.
<!-- This proof of concept provides support for our use of fastText in comparing different languages with varying levels of relatedness.-->

[^1]: With the preregistered exclusion criterion, all three corpora were significant predictors.

Language statistics may incorporate both cultural-specific environmental statistics, that is, the experiential *input* to similarity judgments, and culture-specific senses of similarity, the *conceptualization* of similarity.
For example, cultural features (like farming) can lead to differences in environmental statistics (seeing cows and grass) and these can influence language (talking more about cows and grass).
But other cultural features (like conceiving of similarity thematically) could also cause individuals to talk differently about the same experiences (mentioning what cows eat rather than what other animals cows are like).
Accordingly, our approach examines the extent to which language statistics can predict cross-cultural differences in similarity judgments with the understanding that language statistics are likely a proxy for both *input* to and *conceptualization* of similarity.

@@

```{=html}
<!--@Ji2004 suggest that the notion of similarity is what varies across cultural groups, but it is possible that similarity is regarded in the same way and it is actually the input to this similarity judgment that varies across cultural contexts.
It may be possible to gain traction on potential mechanisms by examining whether variation in similarity judgments co-vary with environmental statistics that differ across cultural and linguistic contexts.--
```
<!-- The present study tests whether the statistics of language can be used to predict cross-cultural differences in similarity judgments, and particularly, whether this approach provides additional insight beyond cross-cultural characterizations based on taxonomic vs thematic preferences. -->

@@

We also applied four demographic exclusion criteria designed to reduce cross-cultural influences across our samples: we excluded participants who (1) were not native speakers of the test language (English, Vietnamese, or Mandarin Chinese), (2) were fluent in another one of the study languages, (3) had lived outside of the test country for more than two years, or (4) had significant international experience (more than 6 international experiences of 2 days or longer).
Following our pre-registration, we dropped any exclusion criterion that would exclude 25% or more of any one population, but only for that population.
Specifically, we did not exclude VN and CN participants who spoke English, or US participants with significant international experience.
As a result of these demographic exclusions, we excluded `r df.excl_US$ppts_after_att_check_excl - df.excl_US$ppts_after_demog_excl` US, `r df.excl_CN$ppts_after_att_check_excl - df.excl_CN$ppts_after_demog_excl` CN, and `r df.excl_VN$ppts_after_att_check_excl - df.excl_VN$ppts_after_demog_excl` VN participants.
The final US sample included `r df.excl_US$ppts_after_demog_excl` participants (`r df.gender_E %>% filter(gender == "Male") %>% pull(n)`M, `r df.gender_E %>% filter(gender == "Female") %>% pull(n)`F, `r df.gender_E %>% filter(gender == "Non-binary") %>% pull(n)` non-binary, `r df.gender_E %>% filter(gender == "Decline to answer") %>% pull(n)` other), with mean age = `r round(df.age_E$mean_age, 2)` (SD = `r round(df.age_E$sd_age, 2)`).
The CN sample included `r df.excl_CN$ppts_after_demog_excl` participants (`r df.gender_M %>% filter(gender == "男性") %>% pull(n)`M, `r df.gender_M %>% filter(gender == "女性") %>% pull(n)`F, `r df.gender_M %>% filter(gender == "拒绝回答") %>% pull(n)` other), with mean age = `r round(df.age_M$mean_age, 2)` (SD = `r round(df.age_M$sd_age, 2)`).
And the VN sample included `r df.excl_VN$ppts_after_demog_excl` participants (`r df.gender_V %>% filter(gender == "Nam") %>% pull(n)`M, `r df.gender_V %>% filter(gender == "Nữ") %>% pull(n)`F, `r df.gender_V %>% filter(gender == "Từ chối trả lời") %>% pull(n)` other), with mean age = `r round(df.age_V$mean_age, 2)` (SD = `r round(df.age_V$sd_age, 2)`).

@@

## Procedure

```{=tex}
\begin{CJK*}{UTF8}{gbsn}
Each participant completed all 105 triads in blocks of 21 trials at a time (10 test triads, 10 filler triads, and 1 attention check per page), by selecting the match most related to the cue. We elicited similarity match responses by asking "Which thing is most closely related to the bolded [cue] item?" In Mandarin Chinese, we used the phrasing for "most closely related" reported by Ji et al. (2004): "哪个事物与加粗项关联最紧密?" And in Vietnamese, we translated this as "Thứ nào liên quan nhất với thứ được in đậm?".
The test triads were presented with 105 filler triads mixed in, to obscure the taxonomic-thematic two-answer forced choice structure of the test stimuli and reduce the likelihood that participants would become aware of the design.
The filler triads were groups of three semantically related words, but where the match options were not distinguished by thematic vs. taxonomic similarity, for example, the cue "bird" with match options "lizard" and "toad." Additionally, we included 11 attention check trials, which were formatted similarly to the test and filler triads but included an instruction instead of a cue item, e.g., "Choose wife" with match options "wife" and "husband." In total, each participant completed 210 similarity judgments (105 test triads and 105 fillers) and 11 attention check questions, with all items presented in randomized orders that varied between subjects.
\end{CJK*}
```
## Corpus model

@@

We found a significant effect of country ($\chi^2$(`r fit.country.anova["country", "Df"]`) = `r round(fit.country.anova["country", "Chisq"], 2)`, p `r ifelse( fit.country.anova["country", "Pr(>Chisq)"] > 0.001, paste0("= ", round(fit.country.anova["country", "Pr(>Chisq)"], 3)), "< .001")`), but this effect is driven by the difference between US and CN responding ($\beta_{US}$ = `r round(fixef(fit.country)["countryUS"], 2)`, p `r ifelse(as.data.frame(summary(fit.country)[["coefficients"]])["countryUS", "Pr(>|z|)"] > 0.001, paste0("= ", round(as.data.frame(summary(fit.country)[["coefficients"]])["countryUS", "Pr(>|z|)"], 3)), "< .001")`). There was no statistical difference between matching in Vietnam and China ($\beta_{Vietnam}$ = `r round(fixef(fit.country)["countryVietnam"], 2)`, p `r ifelse(as.data.frame(summary(fit.country)[["coefficients"]])["countryVietnam", "Pr(>|z|)"] > 0.001, paste0("= ", round(as.data.frame(summary(fit.country)[["coefficients"]])["countryVietnam", "Pr(>|z|)"], 3)), "< .001")`), or the US and Vietnam ($\beta_{Vietnam}$ = `r round(fixef(fit.country_EN_VN)["countryVietnam"], 2)`, p `r ifelse(as.data.frame(summary(fit.country_EN_VN)[["coefficients"]])["countryVietnam", "Pr(>|z|)"] > 0.001, paste0("= ", round(as.data.frame(summary(fit.country_EN_VN)[["coefficients"]])["countryVietnam", "Pr(>|z|)"], 3)), "< .001")`).

@@

For US responding: only the English (EN) corpus was a significant predictor[^2].
EN corpus: $\beta$ = `r round(fixef(fit.ENVIZH_US_cos)["theme_cosine_prop_E"], 2)`, $\chi^2$(`r fit.ENVIZH_US_cos.anova["theme_cosine_prop_E", "Df"]`) = `r round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_E", "Chisq"], 2)`, p `r ifelse( fit.ENVIZH_US_cos.anova["theme_cosine_prop_E", "Pr(>Chisq)"] > 0.001, paste0("= ", round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_E", "Pr(>Chisq)"], 3)), "< .001")`. VI corpus: $\beta$ = `r round(fixef(fit.ENVIZH_US_cos)["theme_cosine_prop_V"], 2)`, $\chi^2$(`r fit.ENVIZH_US_cos.anova["theme_cosine_prop_V", "Df"]`) = `r round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_V", "Chisq"], 2)`, p `r ifelse( fit.ENVIZH_US_cos.anova["theme_cosine_prop_V", "Pr(>Chisq)"] > 0.001, paste0("= ", round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_V", "Pr(>Chisq)"], 3)), "< .001")`. ZH corpus: $\beta$ = `r round(fixef(fit.ENVIZH_US_cos)["theme_cosine_prop_M"], 2)`, $\chi^2$(`r fit.ENVIZH_US_cos.anova["theme_cosine_prop_M", "Df"]`) = `r round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_M", "Chisq"], 2)`, p `r ifelse( fit.ENVIZH_US_cos.anova["theme_cosine_prop_M", "Pr(>Chisq)"] > 0.001, paste0("= ", round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_M", "Pr(>Chisq)"], 3)), "< .001")`.

@@

Additionally, language, culture, cognition, and individual experiences are intertwined in complex causal relationships.
In this study, we measure language and its relation to cross-cultural differences in categorization, but these relations test only the plausibility of a language-based account; they cannot establish the direction of causality.

@Ji2004 established that culture-aligned differences in this paradigm exist, even when the test language is held constant, concluding that "it is culture (independent of the testing language) that led to different grouping styles" in their study. Our data provide a cautionary note to this conclusion, suggesting that while cultural differences in similarity judgment exists, we might get better traction on modelling cross-cultural variation through environmental statistics such as lexical co-occurrences.

There are still many open questions for this account, however. Our operationalization of environmental statistics with word embeddings likely captures both language inputs and culture-specific conceptualization of similarity. Further investigation of the semantic spaces captured by cross-cultural word embedding models is required. Future work should also aim to provide a more specific computational account of how lexical co-occurrence might guide categorization preference beyond the simple proportion-of-similarity model tested here, and investigate other individual- or triad-specific factors that might influence how environmental statistics affect similarity judgment.

>>>>

---
title: "Re-examining cross-cultural similarity judgments using language statistics"
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage[T1, T2A, T5]{fontenc}
  - \usepackage{kotex}
  - \usepackage{CJKutf8}
author-information: > 
    \author{{\large \bf Khuyen N. Le (knl005@ucsd.edu)}$^{1}$, {\large \bf Alexandra Carstensen (abc@ucsd.edu)}$^{1}$, \\ {\large \bf Shan Gao (shangaocog@gmail.com)$^2$}, {\large \bf Michael C. Frank (mcfrank@stanford.edu)$^3$},  \AND
    $^1$Department of Psychology, University of California, San Diego, $^2$Department of Psychology, University of Chicago, \\ $^3$Department of Psychology, Stanford University
    }
abstract: >
 Is “cow” more closely related to “grass” or to “chicken”? Speakers of different languages judge similarity in this context differently, but why? One possibility is that cultures co-varying with these languages induce differences in conceptualizations of similarity. Specifically, East Asian cultures may promote reasoning about thematic similarity, by which cow and grass are more related, whereas Western cultures may bias judgments toward taxonomic relations, like cow and chicken. We measure similarity judgments across the US, China, and Vietnam and replicate US-China differences, but do not find that responding in Vietnam patterns with China. Instead, similarity judgments in Vietnam are intermediate between the US and China. We also show that word embedding models (fastText models for each language) are related to judgments within each country, suggesting a possible alternative interpretation of cross cultural differences. Perhaps notions of similarity are similar across contexts, but the statistics of the linguistic environment vary.

@@

Our study uses cosine distances of fastText word vectors as a measure of lexical co-occurrence.[^2]
fastText is a system that uses lexical co-occurrence information to generate a vector representing each word in its lexicon [@Mikolov2018].
Practically speaking, fastText vectors are available for many languages, and fastText has also been shown to be sensitive to cultural differences in word meanings: @Thompson2020 demonstrated that the distribution of semantic meaning clusters generated by fastText trained on language-specific corpora correlates with the cultural, historical, and geographical similarities of these languages.
<!-- This proof of concept provides support for our use of fastText in comparing different languages with varying levels of relatedness.-->

[^1]: We also carried out our analysis using raw lexical co-occurrences and obtained similar results.

Language statistics may incorporate both cultural-specific environmental statistics, that is, the experiential *input* to similarity judgments, and culture-specific senses of similarity, the *conceptualization* of similarity.
For example, cultural features (like farming) can lead to differences in environmental statistics (seeing cows and grass) and these can influence language (talking more about cows and grass).
But other cultural features (like conceiving of similarity thematically) could also cause individuals to talk differently about the same experiences (mentioning what cows eat rather than what other animals cows are like).
Accordingly, our approach examines the extent to which language statistics can predict cross-cultural differences in similarity judgments with the understanding that language statistics are likely a proxy for both input and conceptualization.

@@

```{=html}
<!--@Ji2004 suggest that the notion of similarity is what varies across cultural groups, but it is possible that similarity is regarded in the same way and it is actually the input to this similarity judgment that varies across cultural contexts.
It may be possible to gain traction on potential mechanisms by examining whether variation in similarity judgments co-vary with environmental statistics that differ across cultural and linguistic contexts.--
```

<!-- The present study tests whether the statistics of language can be used to predict cross-cultural differences in similarity judgments, and particularly, whether this approach provides additional insight beyond cross-cultural characterizations based on taxonomic vs thematic preferences. -->

@@

We also applied four demographic exclusion criteria designed to reduce cross-cultural influences across our samples: we excluded participants who (1) were not native speakers of the test language (English, Vietnamese, or Mandarin Chinese), (2) were fluent in another one of the study languages, (3) had lived outside of the test country for more than two years, or (4) had significant international experience (more than 6 international experiences of 2 days or longer).
Following our pre-registration, we dropped any exclusion criterion that would exclude 25% or more of any one population, but only for that population.
Specifically, we did not exclude VN and CN participants who spoke English, or US participants with significant international experience.
As a result of these demographic exclusions, we excluded `r df.excl_US$ppts_after_att_check_excl - df.excl_US$ppts_after_demog_excl` US, `r df.excl_CN$ppts_after_att_check_excl - df.excl_CN$ppts_after_demog_excl` CN, and `r df.excl_VN$ppts_after_att_check_excl - df.excl_VN$ppts_after_demog_excl` VN participants.
The final US sample included `r df.excl_US$ppts_after_demog_excl` participants (`r df.gender_E %>% filter(gender == "Male") %>% pull(n)`M, `r df.gender_E %>% filter(gender == "Female") %>% pull(n)`F, `r df.gender_E %>% filter(gender == "Non-binary") %>% pull(n)` non-binary, `r df.gender_E %>% filter(gender == "Decline to answer") %>% pull(n)` other), with mean age = `r round(df.age_E$mean_age, 2)` (SD = `r round(df.age_E$sd_age, 2)`).
The CN sample included `r df.excl_CN$ppts_after_demog_excl` participants (`r df.gender_M %>% filter(gender == "男性") %>% pull(n)`M, `r df.gender_M %>% filter(gender == "女性") %>% pull(n)`F, `r df.gender_M %>% filter(gender == "拒绝回答") %>% pull(n)` other), with mean age = `r round(df.age_M$mean_age, 2)` (SD = `r round(df.age_M$sd_age, 2)`).
And the VN sample included `r df.excl_VN$ppts_after_demog_excl` participants (`r df.gender_V %>% filter(gender == "Nam") %>% pull(n)`M, `r df.gender_V %>% filter(gender == "Nữ") %>% pull(n)`F, `r df.gender_V %>% filter(gender == "Từ chối trả lời") %>% pull(n)` other), with mean age = `r round(df.age_V$mean_age, 2)` (SD = `r round(df.age_V$sd_age, 2)`).[^2]

@@

## Procedure
\begin{CJK*}{UTF8}{gbsn}
Each participant completed all 105 triads in blocks of 21 trials at a time (10 test triads, 10 filler triads, and 1 attention check per page), by selecting the match most related to the cue. We elicited similarity match responses by asking "Which thing is most closely related to the bolded [cue] item?" In Mandarin Chinese, we used the phrasing for "most closely related" reported by Ji et al. (2004): "哪个事物与加粗项关联最紧密?" And in Vietnamese, we translated this as "Thứ nào liên quan nhất với thứ được in đậm?".
The test triads were presented with 105 filler triads mixed in, to obscure the taxonomic-thematic two-answer forced choice structure of the test stimuli and reduce the likelihood that participants would become aware of the design.
The filler triads were groups of three semantically related words, but where the match options were not distinguished by thematic vs. taxonomic similarity, for example, the cue "bird" with match options "lizard" and "toad." Additionally, we included 11 attention check trials, which were formatted similarly to the test and filler triads but included an instruction instead of a cue item, e.g., "Choose wife" with match options "wife" and "husband." In total, each participant completed 210 similarity judgments (105 test triads and 105 fillers) and 11 attention check questions, with all items presented in randomized orders that varied between subjects.
\end{CJK*}

## Corpus model

@@

We found a significant effect of country ($\chi^2$(`r fit.country.anova["country", "Df"]`) = `r round(fit.country.anova["country", "Chisq"], 2)`, p `r ifelse( fit.country.anova["country", "Pr(>Chisq)"] > 0.001, paste0("= ", round(fit.country.anova["country", "Pr(>Chisq)"], 3)), "< .001")`), but this effect is driven by the difference between US and CN responding ($\beta$ = `r round(fixef(fit.country)["countryUS"], 2)`, p `r ifelse(as.data.frame(summary(fit.country)[["coefficients"]])["countryUS", "Pr(>|z|)"] > 0.001, paste0("= ", round(as.data.frame(summary(fit.country)[["coefficients"]])["countryUS", "Pr(>|z|)"], 3)), "< .001")`). There was no statistical difference between matching in Vietnam and China ($\beta$ = `r round(fixef(fit.country)["countryVietnam"], 2)`, p `r ifelse(as.data.frame(summary(fit.country)[["coefficients"]])["countryVietnam", "Pr(>|z|)"] > 0.001, paste0("= ", round(as.data.frame(summary(fit.country)[["coefficients"]])["countryVietnam", "Pr(>|z|)"], 3)), "< .001")`), or the US and Vietnam ($\beta$ = `r round(fixef(fit.country_EN_VN)["countryVietnam"], 2)`, p `r ifelse(as.data.frame(summary(fit.country_EN_VN)[["coefficients"]])["countryVietnam", "Pr(>|z|)"] > 0.001, paste0("= ", round(as.data.frame(summary(fit.country_EN_VN)[["coefficients"]])["countryVietnam", "Pr(>|z|)"], 3)), "< .001")`).

@@

For US responding: only the English (EN) corpus was a significant predictor[^5].
EN corpus: $\beta$ = `r round(fixef(fit.ENVIZH_US_cos)["theme_cosine_prop_E"], 2)`, $\chi^2$(`r fit.ENVIZH_US_cos.anova["theme_cosine_prop_E", "Df"]`) = `r round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_E", "Chisq"], 2)`, p `r ifelse( fit.ENVIZH_US_cos.anova["theme_cosine_prop_E", "Pr(>Chisq)"] > 0.001, paste0("= ", round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_E", "Pr(>Chisq)"], 3)), "< .001")`. VI corpus: $\beta$ = `r round(fixef(fit.ENVIZH_US_cos)["theme_cosine_prop_V"], 2)`, $\chi^2$(`r fit.ENVIZH_US_cos.anova["theme_cosine_prop_V", "Df"]`) = `r round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_V", "Chisq"], 2)`, p `r ifelse( fit.ENVIZH_US_cos.anova["theme_cosine_prop_V", "Pr(>Chisq)"] > 0.001, paste0("= ", round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_V", "Pr(>Chisq)"], 3)), "< .001")`. ZH corpus: $\beta$ = `r round(fixef(fit.ENVIZH_US_cos)["theme_cosine_prop_M"], 2)`, $\chi^2$(`r fit.ENVIZH_US_cos.anova["theme_cosine_prop_M", "Df"]`) = `r round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_M", "Chisq"], 2)`, p `r ifelse( fit.ENVIZH_US_cos.anova["theme_cosine_prop_M", "Pr(>Chisq)"] > 0.001, paste0("= ", round(fit.ENVIZH_US_cos.anova["theme_cosine_prop_M", "Pr(>Chisq)"], 3)), "< .001")`.

@@

Additionally, language, culture, cognition, and individual experiences are intertwined in complex causal relationships.
In this study, we measure language and its relation to cross-cultural differences in similarity judgments, but these relations test only the plausibility of a language-based account; they cannot establish the direction of causality.

@Ji2004 established that culture-aligned differences in this paradigm exist, even when the test language is held constant, concluding that "it is culture (independent of the testing language) that led to different grouping styles" in their study. Our data provide a cautionary note to this conclusion, suggesting that while cultural differences in similarity judgments exist, we might get better traction on modelling cross-cultural variation through environmental statistics such as lexical co-occurrence.

There are still many open questions for this account, however. Our operationalization of environmental statistics with word embeddings likely captures both language inputs and culture-specific conceptualization of similarity. Further investigation of the semantic spaces captured by cross-cultural word embedding models is required. Future work should also aim to provide a more specific computational account of how lexical co-occurrence might guide categorization preference beyond the simple proportion-of-similarity model tested here, and investigate other person- or triad-specific factors that might influence how environmental statistics affect similarity judgments.
